---
title: "DS4B-short-report-1, Part 1"
author: "Tsz Ching Li and U7630977"
date: "1/9/2023"
Location on GitHub: https://github.com/sharonli126/DS4B-short-report-1 

output:  
    html_document:
        toc: true
        toc_depth: 4
        theme: cosmo
        number_sections: false
        toc_float: true
        highlight: pygments
        fig_width: 8
        fig_height: 4
---
```{r, include=F}
library(tidyverse)
library(styler)
library(dplyr)
```

# Loading the data
```{r}
data_sea_ice <- read_csv("raw_data/S_seaice_extent_daily_v3.0.csv")
```


# Data Tidy and Summarise
```{r, include=FALSE}
# Changing the string of the data for easier analysation
data_sea_ice$Year <-  as.integer(data_sea_ice$Year)
data_sea_ice$Extent <- as.numeric(data_sea_ice$Extent)
data_sea_ice$Month <-  as.integer(data_sea_ice$Month)
```


```{r}
# Organizing the data i would like to show
average_before_2022 <- data_sea_ice |>
  filter(Year < 2022)|>
  group_by(Month) |>
  summarise(Year = "1978-2021", mean_extent = mean(Extent, na.rm = T)) # the Year here is a character

average_2022to2023 <-  data_sea_ice |>
  filter(Year >= 2022)|>
  group_by(Year, Month) |>
  summarise(mean_extent = mean(Extent, na.rm = T))

# Getting the upper quantile, lower quantile, up and bottom bound from the average in 1978 - 2022
quantile <- data_sea_ice |>
  filter(Year < 2022)|>
  group_by(Month) |>
  summarise(
    upper_quantile = quantile(Extent, 0.75, na.rm = T), 
    lower_quantile = quantile(Extent, 0.25, na.rm = T)
    ) |>
  mutate(IQR = upper_quantile - lower_quantile,
         upper_bound = upper_quantile + 1.5 * IQR,
         lower_bound = lower_quantile - 1.5 * IQR) # this is a formula i took reference from chatGPT 4.0
  
# I have to change "2023" to character show it can combine with average_before_2023 later
average_2022to2023$Year <- as.character(average_2022to2023$Year)

# Combining both the 2023 and 1978-2022 data to use color function later in ggplot
average_ice_extent <-  bind_rows(average_2022to2023, average_before_2022)

# Using letter form of month name (this two codes took reference from chatGPT 4.0)
average_ice_extent$Month <- factor(month.abb[as.numeric(average_ice_extent$Month)], levels = month.abb)

quantile$Month  <- factor(month.abb[as.numeric(quantile$Month)], levels = month.abb)



```


# Data Plotting
```{r, fig.height = 6, fig.width = 10}
# This is a plot I got idea from: https://earthobservatory.nasa.gov/world-of-change/sea-ice-arctic#:~:text=This%20sea%20ice%20grows%20dramatically,become%20negative%20in%20recent%20decades.
# But I have added IQR element to the plot for statiscal support of my statement later

ggplot() +
  geom_point(
    data = average_ice_extent, 
    aes(x = Month, y = mean_extent, color = as.factor(Year)), 
    size = 1
    ) +
  geom_ribbon(
    data = quantile, 
    aes(x = as.numeric(Month), ymin = lower_quantile, ymax = upper_quantile), 
    fill = "grey", 
    alpha = .5
    ) +
  geom_ribbon(
    data = quantile, 
    aes(x = as.numeric(Month), ymin = lower_bound, ymax = upper_bound), 
    fill = "grey", 
    alpha = .2
    ) +
  geom_line(
    data = average_ice_extent, 
    aes(x = as.numeric(Month), y = mean_extent, color = as.factor(Year))
    ) +
  labs(x = "Month",
    y = "Average Sea Ice Extent (millions of km²)",
    title = "Monthly Sea Ice Extent for Each Year in Antarctic Sea", 
    color = "Year") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 16)) 
  
```

# Presnetation script
Here is a plot showing the average sea ice extent for each year in Antarctic Sea. The plot shows 3 curves, including the red curve demonstrating the average of sea ice extent within 1978 to 2021 by month, the green line represents average for 2022, and red line represents the average of 2023 up until August.

Two grey shaded area surrounding the red line is the inner-quantile range for data within 1978 to 2021. The darker grey area includes middle 50% of the data , and the lighter grey area shaded the upper bound and lower bound, observation fell outside of it are considered as outlier, certainly diverse from the average. 

We can obviously see that the average ice extent has already fall lower than the lower quantile in 2022, but still inside the light grey area, entering an uncommon but acceptable range. In contrast, average in 2023 is significantly lower than 2022, especially during the winter season, June to August, which also fells beyond the light area and considered as outlier. This means 2023 is having an abnormally low average ice extent compared to the previous 45 years.

The sea ice extent is a direct barometer for global temperature shift. The aberrations in 2023 is not only an outlier but an indication of a more possibly alarming trend to warn us about the acceleration in global warming pattern, stressing an urgent call from our earth. It is essential to take actions ASAP!


